Introducing an Ecological Trait-data Standard

### 2017-09-25 manuscript/manuscript.Rmd
major edits on manuscript

---
title: "Introducing an Ecological Trait-data Standard"
author: "Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Nadja K. Simons, Andreas Ostrowski, Caterina Penone"
csl: amnat.csl
output:
  word_document:
    keep_md: yes
  pdf_document: default
geometry: margin=1.5in
header-includes: \usepackage{lineno}
fontsize: 12pt
bibliography: div05.bib
---

@@

@@

*< importance of trait based ecology, current research questions, current applications of traits. >*

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07] and have been used to describe the role of species in ecosystems or their ability to persist under certain environmental conditions (@refs).
<!--^[Some frameworks differentiate effect traits and response traits (or performance and fitness traits), depending on the research question [refs, @violle07]]-->
<!--Nadja: I would not mention the distinction between effect and response traits-->
Traits have also proven useful for bridging individual level behaviour and physiology to processes at the ecosystem scale [@diaz13], for understanding physiological and evolutionary trade-offs, and for identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15]. Inferring an unknown trophic role or ecological function of individuals from their apparent features [@duarte11; @fontaine05] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and have not been observed in the ecosystem.
<!--Nadja: I am not sure if I understand what you mean with "have not been observed in the ecosystem".-->
<!--Malte: I guess, you mean their behaviour has not been observed, i.e. might know what they look like (what traits they have) because we caught / found individuals, but we have no idea what they do, where / how they live, with whom they interact etc. How about "... are yet undescribed and their behaviour, their interactions with other organisms and the environment and their ecological function are virtually unknown."-->

On the level of communities, community weighted means or variance of functional traits are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @hopfenmuller14; @jennings09; @moretti13; @bello10; @vandewalle10; @Gossner2016; @Simons2015Ecology; @Gamez-Virues2015; @Birkhofer2014]. Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem (@refs) or of ecosystem services [@wood15; @bello10].
<!--Malte: It is not only CWM's and variance of traits, but also functional diversity metrics that are commonly used to understand change sin functional composition. I would probably add that here.-->

Traits furthermore allow comparison across trophic levels. Just as for pollinator and plant trait-pairs, resource defense traits and predator feeding traits have co-evolved. This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens which may be used to inform ecological networks [@munoz17; @schleuning] and parameterize ecological population-dynamic models [@refs].This is also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups.   
<!--Even at the macroecological scale, adaptions against large herbivore grazing pressure have been shown in global patterns of plant defense traits [@diaz07]. 
Nadja: I am not sure how this fits here as it is very specific. Maybe phrase it more generally?-->
<!--Malte: First sentence, 'comparison across trophic levels". Not sure if this is the ideal wording; is it really comparisons that they allow for here? What about "Functional traits furthermore are an important tool for studying interactions across trophic levels." -->

Traits link ecological research to evolution, management (functions and services) for biodiversity conservation (land use intensification), and climate research (environmental filters). As such, a global synthesis of traits would offer major insights in ecosystem processes and enable projections in future climate and land use scenarios. 
<!--Nadja: Not sure if my changes make sense. I also think that the brackets make this sentence too complicated.-->
Given the expected future challenges in dealing with climate change and managing land use, such global trait synthesis is an interdisciplinary/cross-topic field with great potential.

<!--In recent years, functional ecology has thus grown into a major field in ecological research!
Nadja: We don't really need this sentence.-->

Despite the obvious benefit of such global trait synthesis, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus. A harmonization of trait data standards would be of huge benefit for the advancement of trait-based research and enable efficient data use for future generations.

<!-- 
Flo: This section should probably include some more exploratories papers.

Traits can be measurements of organism morphology, life-history characteristics such as reproductive strategies, physiological traits including metabolism and photosynthetic activity, feeding traits, biochemical and isotopic compounds, behavioural traits, as well as environmental traits.

Refer to Kattge et al 2011, Violle et al 2007, Diaz et al 2007, McGill et al. 2006, Aubin et al. 2013, Lavorel et al. 2002...
## trait-based research within the Biodiversity Exploratories

Refer to successfully published papers on arthropod traits, multi-trophic trait variation etc.

Allan et al 2015
Boerschig et al 2013
Gossner et al 2015 & 2016
Simons et al 2014 & 2016
Kühsel et al 2015 & 2016
Birkhofer et al 2015
Chisté et al 2016
Gamez-Virues et al 2015

Summarize different questions, currently proposed in above- and belowground work within the Exploratories, focus on plant traits, arthropod traits, microbial traits.

Need to unify for multi-trophic integration

Trait matching integrates function across trophic levels, influence of land use and environmental filtering on trait composition, Exploratories are ideal framework for testing

-- Cat: here I would organize a bit differently to link better to the previous section:--
Traits can bridge biodiversity and functioning and that they allow comparision across trophic levels. Furthermore trait composition is influenced by land use and environmental filtering of traits. 
Exploratories are ideal framework for testing these questions because they are framed to study interactions between LUI, biodiversity and functions. 
However, multi-trophic integration is difficult because of different standards, methods, focus..
 needs unification in order to advance research and make data use possible for future generations

<!-- Nadja: I like Caterinas suggestion --> 
<!-- Flo: Yes, makes sense. -->
<!-- Malte: I agree, Caterina's suggestion makes it easier to follow for the reader. -->
<!-- Nadja: I think the section above is now quite close to Caterinas suggestion-->

*< Overview of existing database initiatives. Offering access management and harmonization >*

Trait data have already been standardised into centralised databases for specific organism groups and research questions. Most prominently, plant traits from many different sources --including the extensive LEDA traitbase for the Northwest-European flora [@kleyer08]-- have been collated into the quickly growing TRY database [@kattge11]. TRY collates trait data from other published databases with individually owned datasets, provides an extended access management, data query services and tracking of data usage in studies and secondary publications. Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] or root measurements [Fine-root Ecology Database; @iversen17]. The great achievement of TRY is to harmonize these data into a unified scheme and provide a single platform of global coverage, which takes a huge effort to attract data submissions and map them into a common scheme. Repeating this exercise for other organism groups or other trait-based research questions would be desirable.

Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14]. <!-- more on microbial databases -->

In the animal kingdom, specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems. For instance, The BETSI database collects traits on soil invertebrates (http://betsi.cesab.org/).

<!-- Flo: I would leave this open spreadsheet out for now. It would be quite a lot of effort to collect something representative. It might be an exercise for the exploratories people working on traits, for a start. In principle, any open datasets could be linked and imported into the R-package. I will highlight this possibility when talking about the package. --> 
<!-- As an attempt to track these unstandardized datasets, we initiated a living spreadsheet registry which invites submissions of any open data (Google Spreadsheet link, table 1).

(table of existing databases? name, url, focus, number of records, openness of data, maintained by) -->

<!-- Caterina: I like this table idea but should we limit it to plants and arthropods or also other organisms? Europe or whole world? I would opt for plants and arthropods in Europe but in case this becomes a publication open a collaborative file where anybody in the world can add the information for any organism in any part of the world.

Flo: I like this idea of starting a collaborative list on trait datasets. This
   1. illustrates the hetereogeneity of trait data and the need for a traitdata standard
   2. encourages a global traitdata analysis based on heterogeneous sources, which is enabled by the tools we provide here. See https://github.com/fdschneider/bexis_traits/issues/20
   -->

*< Heterogeneity of data standards and centralised databases >*
<!-- Mandatory Open Access publication still seems to prevent some data providers from uploading their data, why many databases encourage a permissive data policy but do not enforce it. -->
<!-- Malte: I think this definitely needs to be backed up with a reference. -->
<!-- Nadja: I don't think we need to talk about the issue of data not being published openly because we won't fix this just by providing a commont trait data standard -->

<!-- Centralised databases like TRY experience rapid growth but may at some point reach a resource limit in personnel and funding.
Nadja: Not sure where this fits anymore -->

Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades [e.g., @Gossner2015SData,@Gossner2015Ecology,@Renner2014]. It becomes apparent, that different databases and individual datasets come with their own structure, reflecting the research questions of the initiatives behind them and their organismal focus. Additionally, it has been pointed out that data on animals --covering multiple trophic levels and having evolved very different body plans-- will not be a simple task and go far beyond the efforts made for plant trait databases (@refs).

Public research funding agencies increasingly demand the publication of data without access restrictions [@swan12; @allison15], but not all scientific databases for functional traits are compliant with this demand. Additionally, the training on compliance with open data standards within the community of environmental sciences can still be much improved [@schmidt16]. At the same time, the open-access publications in low-threshold data repositories such as figshare, datadryad, researchgate or zenodo are gaining importance and foster a decentralised data hosting with low expectations regarding data standardization and documentation. Given this prospect, there will be no lack of data, but a lack of data standardization hampering efficient future use of published datasets.

<!--Today, several general biodiversity data intitiatives are striving for a global integration of ecological and biological information. Only few so far developed universal data frameworks for trait data, such as the EOL TraitBank, which is the most general framework to date (Parr et al 2015). TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive. -->
To facilitate the standardized upload of research data to reliable and interlinked data infrastructures, data brokering services such as the German Federation on Biological Data (gfbio.org) are likely to gain importance. GFBio is aiming to ease data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data.

Such ontologies are being developed for plant traits, for instance in the TOP, Thesaurus of plant traits, that is used for the TRY database [@garnier17] and the Planteome database and ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org]. On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches or the 'Handbook of protocols for standardized measurement of terrestrial invertebrate functional traits' [@moretti17].

With this paper, we propose a roadmap towards a comprehensive trait data standard, particularly building upon the considerations of @parr16, @kattge11, and @garnier17, and drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).

@@

*< data structure >*

Trait-datasets usually take different formats. For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, they usually are reported in a species $\times$ trait matrix format, with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available. This format is usually reporting missing data as NA. It may store additional information (e.g. on variation of means or literature source) in secondary colums. The matrix format is widely used for the production of lookup-tables at the species level, which for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10]. Also, missing information about the behaviour and functional role of species of little ecological record may be inferred from these tables (@Refs).

Phenotypic variation of traits within species would usually be recorded per observation, i.e. each individual specimen would be recorded in a row, with the respective trait values stored in columns. Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation and sexual dimorphism. We term these tables as occurrence data tables, since each row is centered around a single physical occurrence of a species.<!-- Malte: ... Or a single measurement, right? --> This format is the most intuitive for recording own empirical measurements and therefore is common for measured data and rarely found for reported facts obtained from the literature.

Both the matrix format and the occurrence data format have the different trait values which have been collected or measured on one entity stored in multiple columns. Computationally more effective and allowing for highest flexibility is the storage of traitdata in long-table formats, where each row is reserved for a single measurement or fact of a specific trait, referenced to a single occurrence (i.e. a specimen) via a unique identifier which is assigned to a taxon. This allows for repeated measurements, even on a single individual. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement ID. <!-- Malte: I think the difference between the long-table format and the previous one is not clear enough. Nadja: Is this better now? Malte: Yes, it is. -->
The latter format thus provides the highest resolution and is ideal for storing raw data of trait measurements in huge databases and for linking them via the respective unique identifiers to additional information, such as occurrenceID or locationID.

@@

In this paper, we propose a universal scheme of defined column names and data structure that captures the different degrees of resolution and measurement detail for multiple use cases of trait data. We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.

While existing initiatives for standardising data are focused on a constrained organism group, ecosystem type or region and constitute centralised infrastructures, we aim to facilitate the use of a standard format of trait data by suggessting a flexible formatting scheme that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources. 
In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset, starting with the ones required for our minimal definition of trait-datasets proposed above (Table 1).
We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire record (i.e. metadata). 
Another set of terms is intended to facilitate the definition of traits in ontologies or a thesaurus of traits.

With this framework, we build upon the structure proposed for the Encyclopedia of Life (EOL) TraitBank [@parr16], which uses terms of the Darwin Core standard (DWC). We expand the definitions of these terms for the use case of trait data and add further terms to cover the special demands of trait-based research, while maintaining compatibility. The full list of terms of the traitdata standard including further explanation can be accessed at https://ecologicaltraitdata.github.io/TraitDataStandard/ (will be moved to GFBio Terminology Server before publication).

@@

## minimal requirement of a trait-dataset

@@

The core data are also keeping a record of the scientific taxon for which the measurement or fact was obtained via globally accepted identifiers. To provide an unambiguous reference which is easy to read for researchers and for software, this identifier is provided in the form of a descriptive taxon name (`scientificName`) as well as an unambiguous ID (`taxonID`).

*< global taxonomy standards >*

To enable compatibility across datasets, conventions on trait definitions and taxon names are inevitable. We consider the fact that most authors have their own schemes for standardisation and may contribute to different communities by allowing for a double record of both user-specific and standardised entries (as practiced in TRY, -@kattge11 ). This dupliation of data enables continuity on the author and data owner side and quality checking and comparability on the data user side. Taxonomic standardisation, with respect to ongoing renamings and refinements, is a challenging issue in its own right. We propose to add a second set of columns that contain standardised entries of the taxon name (`scientificNameStd`) which map synonyms to an accepted species name according to a published taxonomic ontology (e.g. GBIF Backbone Terminology) and links to it using a globally unique identifier (`taxonID`). The field `taxonID` therefore should contain a globally accessible URI that maps to the precise taxon definition at the time of the description or measurement (e.g. for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424') . 
Keeping an additional reference to a kingdom and class (using field `kingdom` or `class` as described by Darwin Core Standard) will avoid ambiguities with taxonomic homonyms. 
Additionally, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level. This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15](Shan et al 2012, Schrodt et al 2015).

*< relating data to trait definitions >*

@@

<!-- The initiatives to standardise traits into consensus terminology are usually formed around a specific research question, methodology, or organism group and therefore vary in the traits that they consider important.  -->
Resources for semantic ontologies exist for plant traits (e.g., TOP Plant Trait Ontology [@garnier17]), soil invertebrates (T-Sita, @ref), or biomedical traits of vertebrates (Vertebrate Trait Ontology [@park13]). However, for most organism groups and research contexts we attest a deficit of public ontologies with openly accessible URIs. Therefore, if no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus. In its simplest form, a trait thesaurus would provide a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte12; @garnier17]. 
This listing or table should at minimum provide a human readable, informative trait name (`traitName`) and a unique (alphanumeric) identifier (`traitID`) as well as an unambiguous verbal definition (`traitDescription`), the legit factor levels (for categorical data, `factorLevels`) or expected standard units (for numerical data).

@@

*< other identifiers >*

<!--Additional information may apply to different layers of the core data: on measurement or occurrence, co-variate analysis, environmental filters, spatial distribution. -->

Further detail might be stored along with each measurement or fact reported in the dataset (i.e. each row of the dataset). These details might refer to the origin of the information, the measurement method or statistical precision of the reported value and may be given in a related data sheet which is linked to the data using a unique identifier (`measurementID`) or in additional columns to the core dataset. The terms for this layer of information are defined by the measurement or fact extension. The `measurementID` would also be used to link multivariate measurement data that originate from the same measurement event.

The measurement may also be referred to a single specimen or occurrence via a uniue identifier (`occurrenceID`). For most measured trait data this is a vital information that links related measurements of different traits taken on the same specimen. 
This identifier is usually dataset-specific and can be defined by the author and described with further detail in a separate data table using the columns provided in the occurrence extension detailed below. Some data-types may use global identifiers for occurrence data, e.g. a GBIF URI or a museum collection code that is publicly available.
 <!-- Malte: For readers not that familiar with trait data, the differences between measurement, occurrence and trait ID might not be that clear. It might be worth adding either a sentence explaining in an example what is what OR a tiny figure that shows how a specimen is sampled, processed, traits are measured etc. -->
 <!-- Nadja: I like the idea of having a figure explaining the different terms -->

Additionally, information about the authorship and ownership of the data and the terms of use should be considered. This information may be kept in the metadata of the dataset, but if datasets from different sources are merged, those should be referred to by a unique identifier (`datasetID`) or be reported as additional columns in the merged dataset (`author`, `license`, ...; see Dublin Core Metadata standards, Ref).

@@

<!--

@@

-->

@@

The output of these functions can easily be merged using the `rbind()` function of R base. 
<!-- Malte: Shouldn't this be cbind() ? These functions do not add additional rows but additional columns, right?-->

@@

>>>>

---
title: "Introducing an Ecological Trait-data Standard"
author: Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Nadja K. Simons, Andreas
  Ostrowski, Caterina Penone
csl: amnat.csl
output:
  pdf_document: default
  word_document:
    keep_md: yes
geometry: margin=1.5in
header-includes: \usepackage{lineno}
fontsize: 12pt
bibliography: div05.bib
---

@@

# Abstract

1.	The assessment and analysis of functional traits has been established in all domains of ecological research promising great advancements for biodiversity and climate research. 
2.	Trait data are stored in heterogeneous formats, measured at different resolutions and using different methods, comprising literature and measured data, and focused on a variety of research questions and organism groups. 
3.	To harmonize and standardize trait data, we present a first version of an ecological trait-data standard for storing and sharing functional trait data.
4.	We further provide an R-package that assists in transferring own data structures into the data standard and to harmonize and merge trait datasets from heterogeneous sources. 
5.	With these decentralized tools for traitdata management, we intend to facilitate the exchange and analysis of trait data for ecological research.

@@

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07] and have been used to describe the role of species in ecosystems or their ability to persist under certain environmental conditions (@refs).

Traits have proven useful for bridging individual level physiology and behaviour to processes at the ecosystem scale and patterns of biodiversity at the global level [@diaz13]. Comparative study of traits informs researchers about physiological and evolutionary trade-offs, and helps identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @fontaine05] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. For instance, defense traits and predator feeding traits have co-evolved, just as pollinator and flower traits. This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens which may be used to inform ecological networks [@munoz17; @schleuning] and parameterize ecological population-dynamic models [@refs].

On the level of species assemblages, community weighted means or variance of functional traits or multivariate measures of 'functional diversity' [@petchey??] are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @bello10; @hopfenmuller14; @jennings09; @moretti13;  @vandewalle10; @Gossner2016; @Simons2015Ecology; @Gamez-Virues2015; @Birkhofer2014]. Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem (@refs) or of ecosystem services [@wood15; @bello10].  Due to the abovementioned trait-matching, these analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups.

In recent years, functional ecology has thus grown into a major field of ecological research, linking basic physiological knowledge to functions and services as well as applied matters in land use management and biodiversity conservation, as well as risk assessment of anthropogenic stressors and climate change.  
Given the expected future challenges in dealing with climate change and managing land use, such global trait synthesis is an interdisciplinary/cross-topic field with great potential.

## Overview of trait-data initiatives

Despite the obvious benefit, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus, which renders data on traits heterogeneous. 
Trait data have therefore been standardised into centralised databases for specific organism groups and research questions. These services provide platforms for consensus building in methods, attract data submissions from a focal community of researchers and map them into a common scheme. Furthermore, the databases provide access management services and provide data usage policies, while simplifying data queries for synthesis researchers.

Most prominently, plant traits from many different sources -- including the extensive LEDA traitbase for the Northwest-European flora [@kleyer08] -- have been collated into the quickly growing TRY database [@kattge11].  Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] and or root measurements [Fine-root Ecology Database; @iversen17]. Traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebaur17]. Trait-data for lower plants are less completely assessed. <!-- TRY collates trait data from other published databases with individually owned datasets, provides an extended access management, data query services and tracking of data usage in studies and secondary publications. The great achievement of TRY is to harmonize these data into a unified scheme and provide a single platform of global coverage, which takes a huge effort to attract data submissions and map them into a common scheme. Repeating this exercise for other organism  -->

Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14].<!-- more on microbial databases -->

In the animal kingdom, specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems. For instance, The BETSI database collects traits on soil invertebrates (http://betsi.cesab.org/), or the Carabids.org web portal (http://www.carabids.org/). Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades[e.g., @Gossner2015SData, @Gossner2015Ecology, @Renner2014]. However, a unified platform and harmonizing scheme for the datasets is still lacking. It has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits. [refs]

A global overview of existing trait data for all taxa and trait types is difficult to obtain. As an attempt to track  databases and unstandardized datasets from indipendent publications, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets with their their taxonomic focus, and extend of traits covered, their location on the internet and the terms of use. We invite data owners and users to register further datasets qualifying as trait data.

## Need for ontologies for data standardisation

Public research funding agencies increasingly demand the publication of data without access restrictions [@swan12; @allison15]. However, although the number of trait-based studies grows constantly, not all scientific databases for functional traits are compliant with this demand for open data. <!--The training on compliance with open data standards within the community of environmental sciences can still be much improved [@schmidt16].--> Open-access publications in low-threshold data repositories such as Figshare, datadryad, Researchgate or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardization and documentation. Given this prospect, there will be no lack of data, but a lack of data standardization hampering efficient future use of published datasets.

<!--Today, several general biodiversity data intitiatives are striving for a global integration of ecological and biological information. Only few so far developed universal data frameworks for trait data, such as the EOL TraitBank, which is the most general framework to date (Parr et al 2015). TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive. -->
To facilitate the standardized upload of research data to reliable and interlinked data infrastructures, data brokering services such as the German Federation on Biological Data (gfbio.org) are likely to gain importance. GFBio is aiming to ease data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data.
Such ontologies are being developed for plant traits, for instance in the TOP, Thesaurus of plant traits, that is used for the TRY database [@garnier17] and the Planteome database and ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org]. On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches (T-SITA, http://t-sita.cesab.org/, @pey14).

While these ontologies define traits for focal groups of organisms, they do not specify in which formats or structures trait data should be stored.  Such a general data structure for quantitative trait data has been developed within the Encyclopedia of Life Project for upload of data into TraitBank [@parr16], relying on definitions of terms provided by the Darwin Core Standard (DWC, ...). 
However, the Terms of the Darwin Core Standard and available Extensions do not cover all of the aspects required in the various use-cases of trait-based research.

With this paper, we propose a roadmap towards a comprehensive Ecological Traitdata Standard (EST) for general application in trait-based research, particularly building upon the considerations of @parr16, @kattge11, and @garnier17, and drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).

We propose a versatile data structure, as well as a scheme of defined column names that captures the different degrees of resolution and measurement detail for multiple use cases of trait data. We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.

With this framework, we build upon the structure proposed for the Encyclopedia of Life (EOL) TraitBank [@parr16], which uses terms of the Darwin Core standard (DWC). We expand the definitions of these terms for the use case of trait data and add further terms to cover the special demands of trait-based research, while maintaining compatibility.

@@

Trait-datasets usually take different formats. For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species $\times$ trait matrix format, with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available. This format is usually reporting missing data as NA. It may store additional information (e.g. on variation of means or literature source) in secondary colums. The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon). Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10]. Also, missing information about the behaviour and functional role of species of little ecological record may be inferred from these tables (@Refs).

Phenotypic variation of traits within species would usually be recorded per observation, i.e. each individual specimen would be recorded in a row, with the respective trait values stored in columns. In these 'occurrence data tables', each row is centered around a single physical occurrence of a species. Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation along environmental gradients. Occurence data would also capture phenotypic variation arising from morphotypes and sexual dimorphism. This format is the most intuitive for recording own empirical measurements and therefore is common for measured quantitative data and rarely found for reported qualitative facts obtained from the literature.

Both the matrix format and the occurrence data format have the different trait values which have been collected or measured on one entity stored in multiple columns. Each trait column represents one measured value.

Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single occurrence (i.e. a specimen) via a unique identifier which is assigned to a taxon. This allows for repeated measurements, even on a single individual. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement ID.  Longtable dataset purport multiple advantages for the data manipulation (e.g. filtering, subsetting and aggregating data), visualization (e.g. plot measured values by factor variable or taxon) and statistical modeling (e.g. ANOVA for testing difference of trait value by sex) [@wickham14]. They further draw from a defined set of columns, which makes it easier to merge datasets.

The latter format thus provides the highest resolution and is ideal for storing raw data of trait measurements in huge databases and for linking them via the respective unique identifiers to additional information, such as occurrenceID or locationID. Each row of the dataset can be interpreted as a statement of an 'entity x having a quality y'. To comply with the EST, datasets must be reformatted to the longtable format.

@@

## common terms for trait data

In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset compliant with the EST, starting with the ones required for our minimal definition of 'core' trait-datasets proposed above (Table 1).
We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire record (i.e. metadata). 
Another set of terms is intended to facilitate the definition of traits in ontologies or a thesaurus of traits, consistent with semantic web standards (@ref, see below for further detail).  The full list of terms  can be accessed at https://ecologicaltraitdata.github.io/TraitDataStandard/ (will be moved to GFBio Terminology Server before publication).

A dataset that applies these terms should relate to the EST in its metadata. Since the traitdata standard is a resource under open community development, the metadata also should refer to a version number.  This is best warranted by providing the full URI of the applied  traitdata standard in the metadata (e.g. http://terminologies.gfbio.de/traitdatastandard/v1.0 !invalid URL!).

@@

## core elements of a trait-dataset

@@

The core data are also keeping a record of the scientific taxon for which the measurement or fact was obtained in the form of a descriptive taxon name (`scientificName`).

### relating to global taxonomy standards

Taxonomic standardisation, with respect to ongoing renamings and refinemenets, is a challenging issue in its own right. To enable compatibility across datasets, conventions on trait definitions and taxon names are inevitable. We consider the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards by allowing for a double record of both user-specific and standardised entries [as practiced in TRY; @kattge11]. This dupliation of data enables continuity on the author and data owner side and quality checking and comparability on the data user side. We therefore propose to add a second set of columns that contain standardised entries of the taxon name (`scientificNameStd`) which map synonyms to an accepted species name according to a published taxonomic ontology (e.g. GBIF Backbone Terminology) and links to it using a globally unique identifier (`taxonID`). The field `taxonID` therefore should contain an public,  globally unique Uniform Resource Identifier (URI) that resolves to the precise taxon definition at the time of the description or measurement (e.g. for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424') . 
Keeping an additional reference to a kingdom and class (using field `kingdom` or `class` as described by Darwin Core Standard) will avoid ambiguities with taxonomic homonyms. 
Additionally, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level. This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15](Shan et al 2012, Schrodt et al 2015).

### relating data to trait definitions

@@

<!-- The initiatives to standardise traits into consensus terminology are usually formed around a specific research question, methodology, or organism group and therefore vary in the traits that they consider important.  -->
Resources for semantic ontologies exist for plant traits (e.g., TOP Plant Trait Ontology [@garnier17]), soil invertebrates (T-Sita, @pey14), or biomedical traits of vertebrates [Vertebrate Trait Ontology, @park13]. However, for most organism groups and research contexts we attest a deficit of public ontologies with publicly accessible URIs.

### Defining trait thesauri and ontologies

If no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus. In its simplest form, a trait thesaurus would provide a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte12; @garnier17]. 
This listing or table should at minimum provide a human readable, informative trait name (`traitName`) and define a unique dataset-specific identifier (`traitID`) as well as a short, unambiguous verbal definition (`traitDescription`) which may make use of standard terms provided in other Ontologies. As an Example,  the definition for 'fruit mass' in TOP reads: "the mass (PATO:mass), either fresh or dried, of a fruit (PO:fruit)", referring to Phenotypic Characeristics Ontology PATO and Planteome Plant Ontolgy, PO (http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass). 
The thesaurus should also constrain the legit factor levels (for categorical data, `factorLevels`) or expected standard units (`traitUnitStd` for numerical data). The type of values should be differentiated in the field `valueType` by specifying 'numerical', 'logical', 'integer', 'categorical' traits.

Relating the trait definition to methodological definitions on trait measurements help standardize the physical process of measurements. The Handbook for Measurement of tratits [@moretti17] is a model example of such a methodological approach.

@@

### relating core data to other datasets

Further detail might be stored along with each measurement or fact reported in the dataset (i.e. each row of the dataset). These details might refer to the origin of the information, the measurement method or statistical precision of the reported value and may be given in a related data sheet which is linked to the data using a unique identifier (`measurementID`) or in additional columns to the core dataset. The terms for this layer of information are defined by the 'Measurement or Fact' extension. The `measurementID` would also be used to link multivariate measurement data that originate from the same measurement event.

The measurement may also be referred to a single specimen or occurrence via a uniue identifier (`occurrenceID`). For most measured trait data this is a vital information that links related measurements of different traits taken on the same specimen. 
This identifier is usually dataset-specific and can be defined by the author and described with further detail in a separate data table using the columns provided in the occurrence extension detailed below. Some data-types may benefit from existing global identifiers for occurrence data, e.g. a GBIF URI or a museum collection code referencing the precise specimen at which the measuremetn was taken and that is publicly available.
 <!-- Malte: For readers not that familiar with trait data, the differences between measurement, occurrence and trait ID might not be that clear. It might be worth adding either a sentence explaining in an example what is what OR a tiny figure that shows how a specimen is sampled, processed, traits are measured etc. -->
 <!-- Nadja: I like the idea of having a figure explaining the different terms -->

### Metadata

Additionally, information about the authorship and ownership of the data and the terms of use should be considered.

Individual measurements and facts may already be part of a larger set of traits or a separate trait database before they are added to a combined trait dataset. To retain the rights of the original data contributor, the field `rightsHolder` states the person or organization that owns or manages the rights to the data; `bibliographicCitation` states a bibliographic reference which should be cited when the data is used; and license specifies under which terms and conditions the data can be used, re-used and/or published. This information always applies to one single fact or measurement, further information on the larger dataset which originally contained this entry can be stored in `datasetID`, `datasetName`, `authorLastname` and `authorFirstname`. These columns should hence give credit to the person who compiled the original dataset and signs responsible for the correct identification and reporting of the rights holder.
These information usually may be kept in the metadata of the dataset, but if datasets from different sources are merged, those should be referred to by a unique identifier (`datasetID`) or be reported as additional columns in the merged dataset (`author`, `license`, ...; see Dublin Core Metadata standards, Ref).

@@

@@

@@

Several datasets that have been produced by these functions can easily be merged using the `rbind()` function of R base.

@@

\footnotesize
